---
output:
  html_document:
    css: style.css
    highlight: default
    theme: paper
---

<br>

#### **Course Preparation**

R is a crucial skill for success in this short course. Considering reviewing [R For Beginners](https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf), which gives a nice overview of basic R functionality. Specifically, you should be very comfortable with manipulating matrices and lists and writing and using custom functions.

We strongly recommend purchasing [Hobbs & Hooten 2015](https://www.amazon.com/Bayesian-Models-Statistical-Primer-Ecologists/dp/0691159289). The first three chapters provide foundational material that we will cover fairly quickly in the course, so if you have not had a course in mathematical statistics, reading those chapters before the course is crucial. The structure of the course closely follows the organization of the book and it will be a useful reference after the course.

<br>

#### **MCMCVis R Package**

We will use the `MCMCvis` [package](https://cran.r-project.org/web/packages/MCMCvis/index.html) to manipulate and visualize JAGS output. The [package vignette](https://cran.r-project.org/web/packages/MCMCvis/vignettes/MCMCvis.html) nicely outlines its capabilities. 

<br>

#### **Statistical Notation Help**

Here are some [conventions for statistical notation](content/presentation/ConventionsForNotation.html) that you might find helpful. 

<br>

#### **Install course R package `SESYNCBayes`**

We have prepared an R package that contains all the data you will need to complete our lab exercises.  The package is part of the course materials that you now have on your local machine. You will need to do an initial install of this package and do periodic updates throughout the course. For both the install and update the commands are the same.

1. Download the R package [SESYNCBayes ver. 0.6](content/package/SESYNCBayes_0.6.0.tar.gz) to your computer.

2. Open R or RStudio run the following line of code to install the `SESYNCBayes` package from source. Remember to change `<pathtoSESYNCBayes>` to the path to the directory where you cloned the SESYNCBayes course repository.

``` {r, eval = FALSE, echo = TRUE, include = TRUE}
install.packages("<pathtoSESYNCBayes>/Packages/SESYNCBayes_0.6.0.tar.gz", repos = NULL, type = "source")
```

3. When working in RStudio you load the library like any other R library:

``` {r, eval = FALSE, echo = TRUE, include = TRUE}
library(SESYNCBayes)
```
    
3.  It is easy to see the help files for `SESYNCBayes` type in R:

``` {r, eval = FALSE, echo = TRUE, include = TRUE}
?SESYNCBayes
```
    
<br>

#### **Licensing**

All material on this website is licensed under GPLv3.
